### 2020-12-16 latex/acrobot.tex
Final edit of acrobot draft

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma \simeq \SxR\).
\begin{enumerate}
    \item There exists \(I > 0\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \(\mathcal{O}_1\).
    What's more, almost every orbit will escape the closure of 
    \(\mathcal{O}_1\).
    If instead \(I < 0\), the VNHC dissipates energy.
\item Define \(b : \SxR_{> 0} \rightarrow \R\) by
    \[
        b(\beta,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\beta^2 + 30c_\beta(1 - c_\beta)\right)
            - c_\beta\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\beta)}
        }
        ,
    \]
    and define \(S(\rho_0) := \int \limits_{0}^{2\pi} b(\sigma,\rho_0)d\sigma\).
    Fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose there exists \(\epsilon > 0\) so that \(S(\rho_0) \geq \epsilon\) for
    all \(\rho_0 \in \, \left]\sqrt{60m^2gl^3}, \bar{\rho}\right]\).
    Then there exists \(I > 0\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \(\mathcal{O}_2(\bar{\rho})\).
    If instead \(I < 0\), the VNHC dissipates energy.
\end{enumerate}
\end{thm}

Notice that \(\mathcal{O}_1 \subset \mathcal{O}_2(\bar{\rho})\), yet
Theorem \ref{thm:acrobot-energy-stabilization} considers these sets separately.
This separation is advantageous because the first result holds for any
\(m\), \(g\), \(l\), and \(\bar{q}_a\). 
That is, regardless of one's choice of acrobot, 
a small enough fixed value \(I > 0\) will enable the acrobot to 
escape any compact subset of \(\mathcal{O}_1\) in finite time. 
What's more, the acrobot's phase will flow towards (and through) the homoclinic orbit
corresponding to \(E(\pi,0)\), as in Figure
\ref{fig:acrobot-oscillation-domain}.

@@

To begin, let us analyze the upright equilibrium.
Taking the Jacobian of the constrained dynamics
~\eqref{eqn:acrobot-constrained-dynamics} at \((q_u,p_u) = (\pi,0)\) yields
\[
    J = \begin{bmatrix}
        -\frac{6mgl\bar{q}_aI}{5} & \frac{1 - 2m^2gl^3\bar{q}_a^2 I^2}{5ml^2} \\
        3mgl & mgl\bar{q}_aI
    \end{bmatrix}
    ,
\]
which has characteristic polynomial
\[
    \det\left(\lambda \Id{2} - J\right)
    = \lambda^2 + \frac{mgl\bar{q}_a I}{5} \lambda - 3g
    .
\]
By Descartes' rule of signs, this polynomial has one root with positive real
part. 
The equilibrium \((\pi,0)\) is therefore unstable, so the set \(\Pi^+\) of
initial conditions converging to \((\pi,0)\) is one-dimensional, and hence is
of measure zero in \(\SxR\).

Using \(x(t) := (q_u(t),p_u(t))\) as shorthand, let
\(x(0) \in \mathcal{O}_1\) be a nonzero initial condition of the acrobot.
Suppose the orbit \(x(\R)\) does not cross through \(E_\pi\)
into the rotation zone.
Since the acrobot gains energy on \(\mathcal{O}_1\), the positive limit set of
\(x(t)\) must be \(E_\pi\). 
By the Birkhoff Theorem \cite{birkhoff}, \(x(t)\) converges to an invariant
subset of \(E_\pi\).
From our discussion on pseudo-polar coordinates (in particular the extension
~\eqref{eqn:alpha-dot-boundary} of \(\dot{\alpha}\) to \(\bar{\mathcal{O}}_1\)),
we know that \(\dot{\alpha} \geq 0\) on \(E_\pi\), with equality if and only if 
\(\alpha \in \{0,\pi\}\).
In other words, the only invariant subset of \(E_\pi\) is the upright equilibrium.
Hence, \(x(t)\) must converge to \((\pi,0)\), which means \(x(0) \in \Pi^+\).
Since we know that \(\Pi^+\) is a set of measure zero in \(\SxR\), almost every
other orbit initialized in \(\mathcal{O}_1\) must escape through the boundary
\(E_\pi\) into the rotation domain.

@@

We have proven the first part of Theorem \ref{thm:acrobot-energy-stabilization},
which claims there exists \(I > 0\) small enough that 
~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
\(\mathcal{O}_1\).
Enough energy is injected that orbits will exit the closure of \(\mathcal{O}_1\)
and enter the rotation domain.

Here is a summary of the proof:
\begin{enumerate}
    \item We found pseudo-polar coordinates \((\alpha,\mu)\) adapted to level
        sets of the nominal pendulum's mechanical energy on \(\mathcal{O}_1\).
    \item We showed there exists a value \(I_1\) small enough where 
        \(\dot{\alpha} > 0\) everywhere on \(\mathcal{O}_1\).
    \item We found the time-scaled pseudo-radius \(\hat{\mu}\) and
        expanded it into ~\eqref{eqn:acrobot-muhat-approx} using
        perturbation theory.
    \item Taking this expanded solution, we defined and expanded the
        Poincar\'{e} map \(P_\mathcal{O}\) to analyze the discrete-time system
        ~\eqref{eqn:muhat-discrete}.
    \item We found a value \(I_2 \leq I_1\) making the origin a repeller of 
        ~\eqref{eqn:muhat-discrete}, and a value \(I_3 \leq I_2\) which drives
        \(\mu_n\) towards \(\mu = \pi\).
    \item We showed that the upright equilibrium \((\pi,0)\) is unstable, and
        used this fact to prove that almost every orbit crosses through \(E_\pi\)
        into the rotation zone.
\end{enumerate}
Choosing \(I \in \,]0,I_3]\) guarantees energy injection on
\(\mathcal{O}_1\) because all orbits approach the pseudo-radius \(\mu = \pi\).
Likewise, choosing \(I \in [-I_3,0[\) guarantees energy dissipation on
\(\mathcal{O}_1\).

@@

Here is a summary of the proof:
\begin{enumerate}
    \item We found pseudo-polar coordinates \((\beta,\rho)\) adapted to level
        sets of the nominal pendulum's mechanical energy on \(\mathcal{R}\).
    \item We showed there exists a value \(I_1\) small enough where
        \(\dot{\beta} > 0\) everywhere on \(\mathcal{R}\).
    \item We found the time-scaled pseudo-radius \(\hat{\rho}\) and expanded it
        into ~\eqref{eqn:acrobot-rhohat-approx} using perturbation theory.
    \item Taking this expanded solution, we defined and expanded the
        Poincar\'{e} map \(P_\mathcal{R}\) on the domain 
        \(D = D^- \cup D^+\) which shows how the acrobot's orbits behave on
        \(\mathcal{R}_{\bar{\rho}}\).
    \item Using the Poincar\'{e} map to analyze the discrete-time system
        ~\eqref{eqn:rhohat-discrete}, we found values \(I_2,I_4\) bounding the
        remainder term \(R(\beta,\rho_0,I)\) on \(D^+\) and \(D^-\)
        (respectively), then found values \(I_3, I_5\) which drive all orbits of
        ~\eqref{eqn:rhohat-discrete} towards \(\abs{\rho} = \bar{\rho}\).
\end{enumerate}
Recall from Section \ref{sec:acrobot-proof-o1} that there exists \(I^\star > 0\)
making orbits escape \(\bar{\mathcal{O}}_1\) into 
\(\mathcal{R}\). 
Choosing \(I \in \,]0,\min\{I_3,I_5,I^\star\}]\)
guarantees energy injection on \(\mathcal{O}_2(\bar{\rho})\).
Likewise, choosing \(I \in [-\min\{I_3,I_5,I^\star\},0[\) guarantees energy
dissipation on \(\mathcal{O}_2(\bar{\rho})\).

>>>>

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma \simeq \SxR\).
\begin{enumerate}
    \item There exists \(I^\star > 0\) such that, for all 
        \(I \in \,]0,I^\star]\), 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \(\mathcal{O}_1\).
    Moreover, almost every orbit will escape the closure of 
    \(\mathcal{O}_1\) in finite time.
    If instead \(I \in [-I^\star,0[\), the VNHC dissipates energy.
\item Define \(b : \SxR_{> 0} \rightarrow \R\) by
    \[
        b(\beta,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\beta^2 + 30c_\beta(1 - c_\beta)\right)
            - c_\beta\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\beta)}
        }
        ,
    \]
    and define \(S(\rho_0) := \int \limits_{0}^{2\pi} b(\sigma,\rho_0)d\sigma\).
    Fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose there exists \(\epsilon > 0\) so that \(S(\rho_0) \geq \epsilon\) for
    all \(\rho_0 \in \, \left]\sqrt{60m^2gl^3}, \bar{\rho}\right]\).
    Then there exists \(I^\star > 0\) such that, for all 
    \(I \in \, ]0,I^\star]\)
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \(\mathcal{O}_2(\bar{\rho})\).
    If instead \(I \in [-I^\star,0[\), the VNHC dissipates energy.
\end{enumerate}
\end{thm}

Notice that \(\mathcal{O}_1 \subset \mathcal{O}_2(\bar{\rho})\), yet
Theorem \ref{thm:acrobot-energy-stabilization} considers these sets separately.
This separation is advantageous because the first result holds for any
\(m\), \(g\), \(l\), and \(\bar{q}_a\). 
That is, regardless of one's choice of acrobot, 
a small enough fixed value \(I > 0\) will enable the acrobot to 
escape any compact subset of \(\mathcal{O}_1\) in finite time.
Therefore, the acrobot will display an oscillatory behaviour whereby the
amplitude of oscillation increases with time.
In so doing, the acrobot's orbit will exit in finite time the closure of
\(\mathcal{O}_1\) as in Figure \ref{fig:acrobot-oscillation-domain}.

@@

To begin, let us analyze the upright equilibrium.
Taking the Jacobian of the constrained dynamics
~\eqref{eqn:acrobot-constrained-dynamics} at \((q_u,p_u) = (\pi,0)\) yields
\[
    J = \begin{bmatrix}
        -\frac{6mgl\bar{q}_aI}{5} & \frac{1 - 2m^2gl^3\bar{q}_a^2 I^2}{5ml^2} \\
        3mgl & mgl\bar{q}_aI
    \end{bmatrix}
    ,
\]
which has characteristic polynomial
\[
    \det\left(\lambda \Id{2} - J\right)
    = \lambda^2 + \frac{mgl\bar{q}_a I}{5} \lambda - 3g
    .
\]
By Descartes' rule of signs, this polynomial has one root with positive real
part. 
The equilibrium \((\pi,0)\) is therefore unstable, so the stable manifold \(\Pi^+\) of
initial conditions converging to \((\pi,0)\) is one-dimensional, and hence is
of measure zero in \(\SxR\).

Using \(x(t) := (q_u(t),p_u(t))\) as shorthand, let
\(x(0) \in \mathcal{O}_1\) be a nonzero initial condition of the acrobot.
Suppose the orbit \(x(\R)\) is confined within 
\(\bar{\mathcal{O}}_1\), and does not exit through \(E_\pi\) into the rotation
zone.
Since \(\bar{\mathcal{O}}_1\) is compact, the Birkhoff Theorem \cite{birkhoff}
implies:
\begin{itemize}
    \item The positive limit set \(L_+\) of \(x(t)\) is non-empty, compact, and
        invariant.
    \item The solution \(x(t)\) asymptotically tends to \(L_+\).
\end{itemize}
Since the acrobot gains energy on \(\mathcal{O}_1\), the positive limit set of
\(x(t)\) must be the largest invariant subset of \(E_\pi\). 
From our discussion on pseudo-polar coordinates (in particular the extension
~\eqref{eqn:alpha-dot-boundary} of \(\dot{\alpha}\) to \(\bar{\mathcal{O}}_1\)),
we know that \(\dot{\alpha} \geq 0\) on \(E_\pi\), with equality if and only if 
\(\alpha \in \{0,\pi\}\).
There are two possibilities: either \(E_\pi\) itself is invariant, or the
largest invariant subset of \(E_\pi\) is \(\{(\pi,0)\}\).
To rule out the first possibility, take the derivative of \(E(q_u,p_u)\) at the
\(p_u\)-axis to get
\[
    \dot{E}(0,p_u) = -\frac{g}{5l} \sin(q_a)p_u
    .
\]
This is non-zero everywhere except at the origin, which means it is non-zero on
\(E_\pi\). 
This means \(E_\pi\) is not invariant, so the positive limit set of \(x(t)\)
must be the set \(\{(\pi,0)\}\).
Hence, \(x(t)\) converges to \((\pi,0)\), which means \(x(0) \in \Pi^+\).
Since we know that \(\Pi^+\) is a set of measure zero in \(\SxR\), almost every
other orbit initialized in \(\mathcal{O}_1\) must escape through the boundary
\(E_\pi\) into the rotation domain.

@@

We have proven the first part of Theorem \ref{thm:acrobot-energy-stabilization},
which claims there exists a control value \(I^\star > 0\) such that, for 
\(I \in \, ]0,I^\star]\), ~\eqref{eqn:acrobot-constraint} injects energy into
the acrobot on \(\mathcal{O}_1\).
Enough energy is injected that orbits will exit the closure of \(\mathcal{O}_1\)
and enter the rotation domain.

Here is a summary of the proof:
\begin{enumerate}
    \item We found pseudo-polar coordinates \((\alpha,\mu)\) adapted to level
        sets of the nominal pendulum's mechanical energy on \(\mathcal{O}_1\),
        where \(\alpha \in \Sone\) is a pseudo-angle and \(\mu \in \, ]0,\pi[\)
        is a pseudo-radius.
    \item We showed there exists a value \(I_1\) small enough where 
        \(\dot{\alpha} > 0\) everywhere on \(\mathcal{O}_1\).
    \item Using \(\alpha\) as a time variable, we found the time-scaled
        pseudo-radius \(\hat{\mu}\) and expanded it into
        ~\eqref{eqn:acrobot-muhat-approx} using perturbation theory.
    \item Taking this expanded solution, we defined and expanded the
        Poincar\'{e} map \(P_\mathcal{O}\) to analyze the discrete-time system
        ~\eqref{eqn:muhat-discrete}.
    \item We found a value \(I_2 \leq I_1\) making the origin a repeller of 
        ~\eqref{eqn:muhat-discrete}, and a value \(I_3 \leq I_2\) which drives
        \(\mu_n\) towards \(\mu = \pi\).
    \item We showed that the upright equilibrium \((\pi,0)\) is unstable, and
        used this fact to prove that almost every orbit crosses through
        \(E_\pi\), thereby exiting the closure of \(\mathcal{O}_1\) in finite
        time.
\end{enumerate}
Choosing \(I \in \,]0,I_3]\) guarantees energy injection on
\(\mathcal{O}_1\) because all orbits approach the pseudo-radius \(\mu = \pi\).
In fact, almost all orbits escape \(\bar{\mathcal{O}}_1\).
Likewise, choosing \(I \in [-I_3,0[\) guarantees energy dissipation on
\(\mathcal{O}_1\).

@@

Here is a summary of the proof:
\begin{enumerate}
    \item We found pseudo-polar coordinates \((\beta,\rho)\) adapted to level
        sets of the nominal pendulum's mechanical energy on \(\mathcal{R}\),
        where \(\beta \in \Sone\) is a pseudo-angle and
        \(\rho \in D\) is a pseudo-radius.
    \item We showed there exists a value \(I_1\) small enough where
        \(\dot{\beta} > 0\) everywhere on \(\mathcal{R}\).
    \item Using \(\beta\) as a time variable, we found the evolution of the
        time-scaled pseudo-radius \(\hat{\rho}\) and expanded it into
        ~\eqref{eqn:acrobot-rhohat-approx} using perturbation theory.
    \item Taking this expanded solution, we defined and expanded the
        Poincar\'{e} map \(P_\mathcal{R}\) on the domain 
        \(D = D^- \cup D^+\) which shows how the acrobot's orbits behave on
        \(\mathcal{R}_{\bar{\rho}}\).
    \item Using the Poincar\'{e} map to analyze the discrete-time system
        ~\eqref{eqn:rhohat-discrete}, we found values \(I_2,I_4\) bounding the
        remainder term \(R(\beta,\rho_0,I)\) on \(D^+\) and \(D^-\)
        (respectively), then found values \(I_3, I_5\) which drive all orbits of
        ~\eqref{eqn:rhohat-discrete} towards \(\abs{\rho} = \bar{\rho}\).
\end{enumerate}
Recall from Section \ref{sec:acrobot-proof-o1} that there exists \(I^\star > 0\)
making orbits escape \(\bar{\mathcal{O}}_1\) into 
\(\mathcal{R}\). 
Choosing \(I \in \,]0,\min\{I_3,I_5,I^\star\}]\)
guarantees energy injection on \(\mathcal{O}_2(\bar{\rho})\).
Likewise, choosing \(I \in [-\min\{I_3,I_5,I^\star\},0[\) guarantees energy
dissipation on \(\mathcal{O}_2(\bar{\rho})\).

This analysis, together with that of section \ref{sec:acrobot-proof-o1},
completes the proof of Theorem \ref{thm:acrobot-energy-stabilization}.
